---
title: "Improvement of chromVAR by TMM normalization - real data"
author: "Jiayi Wang"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
format:
  html:
    toc: true
    code-fold: true
    code-tools: true
    toc_float: true

output:
  quarto::quarto_html:
    output_dir: "../docs/"
---

## Dependencies
```{r set up, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
    library(chromVAR)
    library(motifmatchr)
    library(Matrix)
    library(SummarizedExperiment)
    library(BiocParallel)
    library(BSgenome.Hsapiens.UCSC.hg38)
    library(BSgenome.Mmusculus.UCSC.mm10)
    library(ggplot2)
    library(pheatmap)
    library(patchwork)
    library(limma)
    library(stringr)
    library(ggpubr)
    source("utils.R")
})
set.seed(2017)

```

## Load real data
```{r data}
files <- list.files("../data/02_dat", "*.rds", full.names = TRUE)
dat <- lapply(files, readRDS)
name <- unlist(lapply(files, 
    \(.) substr(basename(.), 1, nchar(basename(.)) - 4)))
name[name == "NR3C1"] <- "GCR"
names(dat) <- name
```

## Standard chromVAR workflow

### Deviations on original data
```{r chromVAR, warning=FALSE}
# get motifs
Hmotifs <- getNonRedundantMotifs(format = "PFMatrix", species = "Hsapiens") 
Mmotifs <- getNonRedundantMotifs(format = "PFMatrix", species = "Mmusculus")

# # run chromVAR
# old_devs <- lapply(names(dat), \(x) {
#   d <- dat[[x]]
#   seqlevelsStyle(d) <- "UCSC"
#   tryCatch(
#   {
#     .chromVAR_bulk(d, Hmotifs, BSgenome.Hsapiens.UCSC.hg38)
#   },
#   error = function(e) {
#     cat("trying BSgenome.Mmusculus.UCSC.mm10...\n")
#     .chromVAR_bulk(d, Mmotifs, BSgenome.Mmusculus.UCSC.mm10)
#   })
# })
# 
# names(old_devs) <- name
# saveRDS(old_devs, "../result/02_dat/old_devs.rds")
old_devs <- readRDS("../result/02_dat/old_devs.rds")
```

### Deviations on Low-accessible regions deviations
```{r low accessbile devs}
# devs_thr50 <- lapply(dat, \(x) {
#     seqlevelsStyle(x) <- "UCSC"
#     rm <- apply(counts(x), 1, min)
#     thr <- quantile(rm)[3]
#     if (thr == 0) thr = 1
#     idx <- which(rm <= thr)
#     x <- x[idx, ]
#     tryCatch(
#     {
#       .chromVAR_bulk(x,
#         Hmotifs,
#         BSgenome.Hsapiens.UCSC.hg38)
#     },
#     error = function(e) {
#       cat("trying BSgenome.Mmusculus.UCSC.mm10...\n")
#       .chromVAR_bulk(x,
#         Mmotifs,
#         BSgenome.Mmusculus.UCSC.mm10)
#     })
# })
# saveRDS(devs_thr50, "../result/02_dat/devs_thr50.rds")
devs_thr50 <- readRDS("../result/02_dat/devs_thr50.rds")
names(devs_thr50) <- name
```


### Deviations on NF fragments
```{r}
# dir <- list.dirs("../data/02a_nucFree", recursive = FALSE)
# nucFree <- lapply(dir, \(.) readRDS(paste0(.,
#   "/runATAC_results/others/countmatrix.rds")))
# names(nucFree) <- basename(dir)
# nucFree_devs <- lapply(names(nucFree), \(x) {
#   d <- dat[[x]]
#   seqlevelsStyle(d) <- "UCSC"
#   message(x)
#   tryCatch(
#   {
#     .chromVAR_bulk(d, 
#       Hmotifs, 
#       BSgenome.Hsapiens.UCSC.hg38)
#   },
#   error = function(e) {
#     cat("trying BSgenome.Mmusculus.UCSC.mm10...\n")
#     .chromVAR_bulk(d, 
#       Mmotifs, 
#       BSgenome.Mmusculus.UCSC.mm10)
#   })
# })
# 
# names(nucFree_devs) <- names(nucFree) 
# saveRDS(nucFree_devs, "../result/02_dat/nucFree_devs.rds")
nucFree_devs <- readRDS("../result/02_dat/nucFree_devs.rds")
names(nucFree_devs)[names(nucFree_devs) == "NR3C1"] <- "GCR"
```

### Deviations on all fragments of nucleosome-free peaks
```{r}
# dir <- list.dirs("../data/02b_fullOnNucFreePeaks/", recursive = FALSE)
# fullNF <- lapply(dir, \(.) readRDS(paste0(.,
#   "/runATAC_results/others/countmatrix.rds")))
# names(fullNF) <- basename(dir)
# fullNF_devs <- lapply(names(fullNF), \(x) {
#   d <- dat[[x]]
#   seqlevelsStyle(d) <- "UCSC"
#   message(x)
#   tryCatch(
#   {
#     .chromVAR_bulk(d, Hmotifs, BSgenome.Hsapiens.UCSC.hg38)
#   },
#   error = function(e) {
#     cat("trying BSgenome.Mmusculus.UCSC.mm10...\n")
#     .chromVAR_bulk(d, Mmotifs, BSgenome.Mmusculus.UCSC.mm10)
#   })
# })
# 
# names(fullNF_devs) <- names(fullNF) 
# saveRDS(fullNF_devs, "../result/02_dat/fullNF_devs.rds")
fullNF_devs <- readRDS("../result/02_dat/fullNF_devs.rds")
names(fullNF_devs)[names(fullNF_devs) == "NR3C1"] <- "GCR"
```


### New method with TMM normalization
```{r new chromVAR}
# new_devs <- lapply(names(dat), \(x) {
#   d <- dat[[x]]
#   seqlevelsStyle(d) <- "UCSC"
#   message(x)
#   tryCatch(
#   {
#     .chromVAR_bulk(d, Hmotifs, BSgenome.Hsapiens.UCSC.hg38, TMM = TRUE)
#   },
#   error = function(e) {
#     cat("trying BSgenome.Mmusculus.UCSC.mm10...\n")
#     .chromVAR_bulk(d, Mmotifs, BSgenome.Mmusculus.UCSC.mm10, TMM = TRUE)
#   })
# })
# names(new_devs) <- name
# saveRDS(new_devs, "../result/02_dat/new_devs.rds")
new_devs <- readRDS("../result/02_dat/new_devs.rds")
```

## Differential analysis of deviations between conditions
```{r new limma, warning=FALSE, message=FALSE}
applyLimma <- \(x) {
  res <- lapply(x, \(dev) {
    colData(dev)$group_id <- substr(colnames(dev), 1, nchar(colnames(dev)) - 5)
    design <- model.matrix(~ dev$group_id)
    fit <- eBayes(lmFit(assays(dev)$z, design))
    res <- topTable(fit, n = Inf)
    ids <- match(rownames(res), rownames(dev))
    res$name <- rowData(dev)$name[ids]
    res$rank <- rank(res$P.Value)
    res
  })
  names(res) <- names(x)
  return(res)
}
old_limma <- applyLimma(old_devs)
low_limma <- applyLimma(devs_thr50)
nucFree_limma <- applyLimma(nucFree_devs)
fullNF_limma <- applyLimma(fullNF_devs)
TMM_limma <- applyLimma(new_devs)
```

## Result evaluation and visualization
### t-statistics

::: {.panel-tabset}
### Original chromVAR
```{r chromVAR t, fig.height=10, fig.width=15}
lapply(names(old_limma), \(x) {
    res <- old_limma[[x]]
    r <- res[res$name == x, "t"]
    ggplot(res, aes(t)) +
      geom_density() + 
      ggtitle(x) +
      geom_vline(xintercept = r, linetype="dashed", 
                  color = "darkred", linewidth=0.8, alpha = 0.8) +
      annotate("text", x = r, y = 0.15, label = round(r,3), 
        color = "darkred", size = 5)
})  |>  wrap_plots(ncol = 4) 
```

### TMM normalization
```{r TMM t, fig.height=10, fig.width=15}
lapply(names(TMM_limma), \(x) {
    res <- TMM_limma[[x]]
    r <- res[res$name == x, "t"]
    ggplot(res, aes(t)) +
      geom_density() + 
      ggtitle(x) +
      geom_vline(xintercept = r, linetype="dashed", 
                  color = "darkred", linewidth=0.8, alpha = 0.8) +
      annotate("text", x = r, y = 0.15, label = round(r,3), 
        color = "darkred", size = 5)
})  |>  wrap_plots(ncol = 4) 
```

### NF fragments
```{r nucFree t, fig.height=10, fig.width=15}
lapply(names(nucFree_limma), \(x) {
    res <- nucFree_limma[[x]]
    r <- res[res$name == x, "t"]
    ggplot(res, aes(t)) +
      geom_density() + 
      ggtitle(x) +
      geom_vline(xintercept = r, linetype="dashed", 
                  color = "darkred", linewidth=0.8, alpha = 0.8) +
      annotate("text", x = r, y = 0.15, label = round(r,3), 
        color = "darkred", size = 5)
})  |>  wrap_plots(ncol = 4) 
```

### All fragments of NF peaks
```{r allF t, fig.height=10, fig.width=15}
lapply(names(fullNF_limma), \(x) {
    res <- fullNF_limma[[x]]
    r <- res[res$name == x, "t"]
    ggplot(res, aes(t)) +
      geom_density() + 
      ggtitle(x) +
      geom_vline(xintercept = r, linetype="dashed", 
                  color = "darkred", linewidth=0.8, alpha = 0.8) +
      annotate("text", x = r, y = 0.15, label = round(r,3), 
        color = "darkred", size = 5)
})  |>  wrap_plots(ncol = 4) 
```

### Low accessible regions
```{r lowAccessible t, fig.height=10, fig.width=15}
lapply(names(low_limma), \(x) {
    res <- low_limma[[x]]
    r <- res[res$name == x, "t"]
    ggplot(res, aes(t)) +
      geom_density() + 
      ggtitle(x) +
      geom_vline(xintercept = r, linetype="dashed", 
                  color = "darkred", linewidth=0.8, alpha = 0.8) +
      annotate("text", x = r, y = 0.15, label = round(r,3), 
        color = "darkred", size = 5)
})  |>  wrap_plots(ncol = 4) 
```

:::

### Distribution of z-scores

::: {.panel-tabset}

### Old: standard chromVAR
```{r z-dist old, fig.height=15, fig.width=10}
lapply(names(old_devs), \(x) {
    dev <- old_devs[[x]]
    ggplot(sechm::meltSE(dev, row.names(dev)), aes(z, colour = sample)) +
      geom_density() + 
      ggtitle(x) +
      geom_vline(xintercept = 0, linetype="dashed", 
                  color = "black", size=0.5, alpha = 0.5)
})  |>  wrap_plots(ncol = 2) 

```

### New: TMM normalization
```{r z-dist new, fig.height=15, fig.width=10}
lapply(names(new_devs), \(x) {
    dev <- new_devs[[x]]
    ggplot(sechm::meltSE(dev, row.names(dev)), aes(z, colour = sample)) +
      geom_density() + ggtitle(x) +
      geom_vline(xintercept = 0, linetype="dashed", 
                  color = "black", size=0.5, alpha = 0.5)
})  |>  wrap_plots(ncol = 2) 

```

### Nucleosome-free fragments
```{r z-dist nucFree, fig.height=15, fig.width=10}
lapply(names(nucFree_devs), \(x) {
    dev <- nucFree_devs[[x]]
    ggplot(sechm::meltSE(dev, row.names(dev)), aes(z, colour = sample)) +
      geom_density() + ggtitle(x) +
      geom_vline(xintercept = 0, linetype="dashed", 
                  color = "black", size=0.5, alpha = 0.5)
})  |>  wrap_plots(ncol = 2) 
```

### All fragments of nucleosome-free peaks
```{r z-dist fullNF, fig.height=15, fig.width=10}
lapply(names(fullNF_devs), \(x) {
    dev <- fullNF_devs[[x]]
    ggplot(sechm::meltSE(dev, row.names(dev)), aes(z, colour = sample)) +
        geom_density() + ggtitle(x) +
      geom_vline(xintercept = 0, linetype="dashed", 
                  color = "black", size=0.5, alpha = 0.5)
})  |>  wrap_plots(ncol = 2) 
```

### Low accessible regions
```{r z-dist low, fig.height=15, fig.width=10}
lapply(names(devs_thr50), \(x) {
    dev <- devs_thr50[[x]]
    ggplot(sechm::meltSE(dev, row.names(dev)), aes(z, colour = sample)) +
        geom_density() + ggtitle(x) +
      geom_vline(xintercept = 0, linetype="dashed", 
                  color = "black", size=0.5, alpha = 0.5)
})  |>  wrap_plots(ncol = 2) 
```

:::

### logFC of limma results

```{r lfc, fig.height=6, fig.width=12}
# lfc <- lapply(names(old_limma), \(x) {
#     old <- data.frame(old_limma[[x]], motif = x, method = "old")
#     new <- data.frame(new_limma[[x]], motif = x, method = "new")
#     rbind(old, new)
# })
# lfc <- do.call(rbind, lfc)
```

::: {.panel-tabset}
### Histogram
```{r lfc histo, fig.height=15, fig.width=10}
# ggplot(lfc, aes(x = logFC, col = method)) + 
#     geom_histogram(aes(y = after_stat(density)),      
#         fill = "white",
#         bins = 100,
#         alpha = 0.2) +
#     geom_density() + 
#     facet_grid(motif ~ method, scales = "free")
```

### Violin plot
```{r violin}
# ggplot(lfc, aes(x = factor(motif), y = logFC, fill = method)) +
#   geom_violin(alpha = 0.5) +
#   geom_boxplot(width = 0.2, position = position_dodge(0.9), alpha = 0.4) +
#   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

::: 

### MA plot

::: {.panel-tabset}

### New method
```{r MA new, fig.height=15, fig.width=20}
# colnames(lfc)[c(1,2,5)] <- c("log2FoldChange", "baseMeanLog2", "padj")
# nlfc <- lfc[lfc$method == "new",]
# olfc <- lfc[lfc$method == "old",]
# lapply(unique(nlfc$motif), \(x) {
#   df <- nlfc[nlfc$motif == x,]
#   ggmaplot(df, dr = 0.05, fc = 2, size = 0.4,
#      palette = c("#B31B21", "#1465AC", "darkgray"),
#      genenames = as.vector(df$name),
#      legend = "top", top = 20,
#      font.label = c("bold", 11),
#      font.legend = "bold",
#      font.main = "bold",
#      ggtheme = ggplot2::theme_minimal()) + ggtitle(x)
# })  |>  wrap_plots(ncol = 4) 

```

### Old
```{r MA old, fig.height=15, fig.width=20}
# lapply(unique(olfc$motif), \(x) {
#   df <- olfc[olfc$motif == x,]
#   ggmaplot(df, dr = 0.05, fc = 2, size = 0.4,
#      palette = c("#B31B21", "#1465AC", "darkgray"),
#      genenames = as.vector(df$name),
#      legend = "top", top = 20,
#      font.label = c("bold", 11),
#      font.legend = "bold",
#      font.main = "bold",
#      ggtheme = ggplot2::theme_minimal()) + ggtitle(x)
# })  |>  wrap_plots(ncol = 4) 
```


:::